---
title: "Reproducible Research Project 1"
author: "Michael Galarnyk"
date: "4/29/2017"
output:
  md_document:
    variant: markdown_github
---

github repo for rest of specialization: [Data Science Coursera](https://github.com/mGalarnyk/datasciencecoursera)

## 1 - Project Title: US Storm Event Human Health and Economic Impact Analysis

## 2 - Synopsis:
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

information on the data: [Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

## 3 - Data Processing:

### 3.1 - Data Loading:
Download the raw data file and extract the data into a dataframe.

```{r}
library("data.table")
library(ggplot2)

fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(fileUrl, destfile = paste0("C:/Users/mgalarnyk/Desktop", '/repdata%2Fdata%2FStormData.csv.bz2'))
stormDF <- read.csv("C:/Users/mgalarnyk/Desktop/repdata%2Fdata%2FStormData.csv.bz2")

stormDT <- as.data.table(stormDF)
```

Explore what is contained in the dataset

```{r}
colnames(stormDT)

str(stormDT)
```

### 3.2 - Data Subsetting:
Subset the dataset on the parameters of interest.
```{r}

# Change parameter names to lowercase.
data.table::setnames(x = stormDT, old = colnames(stormDT), new = tolower(colnames(stormDT)))

# Subset on the parameters of interest.
stormDT <- stormDT[(evtype != "?" & 
             (injuries > 0 | fatalities > 0 | propdmg > 0 | cropdmg > 0)), c("evtype", 
                                                                              "fatalities", 
                                                                              "injuries", 
                                                                              "propdmg", 
                                                                              "propdmgexp", 
                                                                              "cropdmg", 
                                                                              "cropdmgexp") ]
```

### 3.3 - Data Cleansing:

```{r}

# Change all damage exponents to uppercase.
stormDT[, propdmgexp := toupper(propdmgexp)]
stormDT[, cropdmgexp := toupper(cropdmgexp)]

# Map property damage alphanumeric exponents to numeric values.
propDmgKey <-  c("\"\"" = 10^0,
                 "-" = 10^0, 
                 "+" = 10^0,
                 "0" = 10^0,
                 "1" = 10^1,
                 "2" = 10^2,
                 "3" = 10^3,
                 "4" = 10^4,
                 "5" = 10^5,
                 "6" = 10^6,
                 "7" = 10^7,
                 "8" = 10^8,
                 "9" = 10^9,
                 "H" = 10^2,
                 "K" = 10^3,
                 "M" = 10^6,
                 "B" = 10^9)


stormDT[,propdmgexp := propDmgKey[as.character(stormDT[,propdmgexp])]]
stormDT[is.na(propdmgexp), propdmgexp := 10^0 ]

# Map crop damage alphanumeric exponents to numeric values
cropDmgKey <-  c("\"\"" = 10^0,
                "?" = 10^0, 
                "0" = 10^0,
                "K" = 10^3,
                "M" = 10^6,
                "B" = 10^9)
stormDT[, cropdmgexp := cropDmgKey[as.character(stormDT[,cropdmgexp])] ]
stormDT[is.na(cropdmgexp), cropdmgexp := 10^0 ]
```

### 3.4 - Human Heath Data Processing:

```{r}
# Total injuries

totalInjuriesDT <- stormDT[, .(fatalities = sum(fatalities), injuries = sum(injuries), totals = sum(fatalities) + sum(injuries)), by = .(evtype)]


totalInjuriesDT <- totalInjuriesDT[order(-totals), ]

head(totalInjuriesDT,20)
```


## 1 - Project Title: US Storm Event Human Health and Economic Impact Analysis

## 2 - Synopsis:

## 3 - Data Processing:

### 3.1 - Data Loading:

### 3.2 - Data Subsetting:

### 3.3 - Data Cleansing:

### 3.4 - Human Heath Data Processing:

### 3.5 - Economic Data Processing:

## 4 - Analysis Results:

### Figure 4.1: Top 10 Health Impact Event Types

### Figure 4.2: Top 10 Economic Impact Event Types

## 5 - Conclusions

### 5.1 - Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

### 5.2 - Across the United States, which types of events have the greatest economic consequences?

